\chapter*{Acknowledgements}

First and foremost, I could not have completed my PhD without the sponsorship of Soukendai and NICTA.
I’d also like to thank the following individuals:

\begin{itemize}
\item \textbf{Justin Zobel} -- You helped me secure funding to start my PhD, and introduced me to the de Bruijn graph and DNA assembly. Without your help, I may not have started a PhD. You also taught many of the professors I had in University, so I owe a lot to you transitively speaking, too.
\item \textbf{Kathryn Holt} -- Thank you for spending countless hours teaching me about biology in your office. Coming from a school that taught creationism, this was mind expanding, and incredibly influential on the way I would think in general.
\item \textbf{Tom Conway} -- You kicked off the succinct de Bruijn graph subfield, and took the time to help me understand your paper over several lunches.
\item \textbf{Simon Puglisi} -- Thank you for introducing me to succinct data structures. I think I’ve always done my best work whenever I’m with you, whether that is writing a paper, or rolling a cigarette and having a glass of wine in Helsinki.
\item \textbf{Kunihiko Sadakane} -- Thank you for helping a young student pursue his dream of doing a PhD in succinct data structures in Japan, and being patient when I spent too much time having fun in Tokyo. You have encyclopedic knowledge, and I was incredibly lucky to be your student.
\item \textbf{Uno Takeaki} -- You were always able to calm my anxieties of being a 30-something PhD student, and encouraged me to enjoy writing my dissertation. Thank you for your guidance, patience, and understanding.
\item \textbf{Martin Muggli} -- Thank you for being somebody I could ask questions that I really should already know the answer to, and for carrying the codebase on. Thanks as well for the real-talk. I learned a lot about myself from you.
\item \textbf{Travis Gagie, Jouni Siren, and Christina Boucher} -- co-authors, incredibly smart and creative, and a pleasure to work with.
\item \textbf{James Harland} -- For introducing me to research, encouraging exploration, and for being a fantastic mentor.
\item \textbf{Rayan Chikhi, Dinghua Li, Sean Jackman, Jared Simpson, Roberto Grossi, Rajeev Raman, and Srinivasa Rao Satti} -- For being available to answer my questions about a field that was new to me, and for making me feel like a welcome member of our research community.
\item \textbf{Nick Greenfield and Nava Whiteford} -- For hiring me as a succinct de Bruijn graph consultant!
\item \textbf{James Hayton and Javier Grande} -- This dissertation has been a monkey on my back for years, but you cut through my anxieties and helped me get it done.
\end{itemize}

\noindent
Of my family and friends, these people helped me immeasurably:

\begin{itemize}
\item \textbf{John Bowe} -- For finding out exactly how cheap Tokyo is not. Thanks for sponsoring me.
\item \textbf{Christine Mulvey} -- Thank you for helping me learn to program as a kid -- I remember you debugging my basic programs that I typed in from library books. And for doing my assignments in school -- why didn’t you do my dissertation though!?
\item \textbf{James and Nikolas Bowe} -- For the brotherly encouragement and competition. I can finally say I’m better than you both.
\item \textbf{Christian Sea Jones} -- Thanks for looking after me in Tokyo, as well as laughing every time I said “I’m trying to finish my dissertation”. Do or do not, there is no try.
\item \textbf{Ellen Chou} -- For being the first person I met in Tokyo, showing me the cool bars and how to do purikura, and giving me an amazing birthday present a few days after I moved there.
\end{itemize}

\noindent
It was incredibly nostalgic to sit here and write this list and reflect on how grateful I am to have met each of you.
